\documentclass[12pt, twoside]{article}
\usepackage[left=28mm, top=24mm, right=28mm, bottom=24mm, asymmetric, reversemarginpar]{geometry}
\title{\textbf{Disclaimer and Acknowledgements}}
\begin{document}


% Disclaimer
\section{Disclaimer}

I ask that you respect all of the guidelines I set forth with respect to these review notes. If you are unable to respect any of the guidelines mentioned in the proceeding few paragraphs. \\
\begin{itemize}
\item I alone reserve the right to alter and distribute these review notes. If you see a mistake or feel as if important content is missing, contact me (information provided below) so that I can make the appropriate correction(s).
\item I will be supporting the accuracy of these review notes for the remainder of the semester I am taking the course in. After such a time these review notes should be considered obsolete.
\item These review notes are provided absolutely free of charge.
\end{itemize}

Keep in mind that these notes are in no way guaranteed to be accurate. There may be mistakes, outdated information or tangents that will not be directly related to some of the material presented in the course. These notes have been developed by me, a CS undergraduage student who is by no means perfect. No professor, TA or anyone involved in the administration at the University of Waterloo has endorsed these notes.\\

If you feel a need to contact me at any time for clarification, issues with the notes whatever, you can reach me at my contact information below:\\
Jacob Willemsma, (\textbf{wjwillem@uwaterloo.ca}).\\

\section{Acknowledgements}

Some acknowledgments for the content that is found in my course notes.

\begin{itemize}
\item Design inspired by "Linear Algebra Course Notes" (MATH136) by Dan Wolczuk (with permission).
\item CS136 Course Notes by Jacob Pollack and Jacob Willemsma.
\end{itemize}
\clearpage
\thispagestyle{empty}

\end{document}