\section*{Disclaimer}
\addcontentsline{toc}{subsubsection}{Disclaimer}

I ask that you respect all of the guidelines I set forth with respect to these review notes. If you are unable to respect any of the guidelines mentioned in the proceeding few paragraphs, please delete this file or close the hard copy of the review notes.\\

\begin{enumerate}

\item I alone reserve the right to alter and distribute these review notes. If you see a mistake or feel as if important content is missing, contact me (information provided below) so that I can make the appropriate correction(s).

\item I will be supporting the accuracy of these review notes for the remainder of the W14 semester. After such a time these review notes should be considered obsolete.

\item These review notes are provided absolutely free of charge. If you paid for a hard copy or e-copy of these review notes then you are obligated to contact us so that we can take the appropriate action(s) towards the distributor.

\end{enumerate}

Keep in mind that these notes are in no way guaranteed to be accurate. There may be mistakes, outdated information or tangents that will not be directly related to some of the material presented in the course. These notes have been developed by students during our undergraduate year taking CS246. No professor, TA or anyone involved in the administration at the University of Waterloo has endorsed these notes.\\

If you feel a need to contact me at any time for clarification, issues with the notes whatever, you can reach me at my contact information below:\\
Jacob Willemsma, author (\textbf{wjwillem@uwaterloo.ca}).\\

\subsubsection*{Acknowledgements}

Some further acknowledgments to the CS246 material and other influencing figures.

\begin{itemize}

\item Professors and Teaching Assistants for the summer semester of 2014: Brad Lushman and Kirsten Bradley.
\item Design inspired by "Linear Algebra Course Notes" by Dan Wolczuk (with permission).
\item CS246 Course Notes F13 written by Liam Horne.

\end{itemize}